A Comparison of Neural Network Models for Water Quality Forecasting

### 2024-06-11 manuscript/manuscript.Rmd
Headground on results sect

There are many promising neural network architectures that have not been evaluated for water quality forecasting. Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network which was published in 1997 \cite{lstm}. Over the nearly 30 years since LSTM's were introduced, researchers have identified that LSTM's have notable shortcomings, and there have been continued efforts towards the development of better performing models. Many recent studies that use neural networks for water quality forecasting have not explored more contemporary neural network architectures. We will address this gap in the literature by comparing 8 neural network models including LSTM's as well as more recently published neural network models.

@@

We evaluated the forecasting performance of 11 different models on water temperature, dissolved oxygen and chlorophyll-A time series, which were collected by in-situ sensors at 34 freshwater sites across North America. These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON. There is variability across sites in which target variables were observed. Additionally, maintenance issues led to gaps in the data which also varied across locations. Provided the lack of time series and large gaps at certain sites, we evaluated the forecast performance for water temperature, dissolved oxygen and chlorophyll-A at 33, 32 and 9 sites, respectively.

The imputation of missing data proved to be critical to the performance of the ML models. After experimenting with data filling methods that resulted in poor model performance, we developed an imputation method inspired by the climatology model. If the gap size was less than 5 days, then the gap was filled using a Gaussian Process Filter. For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, for the case when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.

@@

where $\nabla$ is the difference operator, $Z_t$ is defined as a theta line and $X_t$ is the original data. When $\theta < 1$, the second differences are reduced from the data, yielding a Theta line that amplifies the long term trends of the data. For instance, if $\theta = 0$, the theta line will be a linear regression of the data. And if $\theta > 1$, the data's short term behavior will be magnified in the Theta line. The Theta model generates a forecast by extrapolating the linear combination of two or more Theta lines. Although the Theta model is relatively simple, it has performed very well in the M3 and M4 forecasting competitions, which has motivated its use in this study as a benchmark statistical model \cite{}. In the time since the Theta model was originally presented in 2000, new variations of the Theta model have been presented which have been shown to outperform the original Theta model \cite{}. Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing variation of the Theta model.

@@

For the 8 neural network models that were investigated, there were varying design choices made regarding which type of covariates could be used. For instance, models could accept past covariates only, future covariates only or both future and past covariates.

For the models that accept past covariates (**list these models**), we used the other target variables recorded at that site as well as air temperature as covariates. For the models that only accept future covariates (**list these models**), we used the day of the year as the covariate. The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.

We provide fully reproducible code for fitting, scoring and visualizing the forecasts. All the machine learning forecasts were generated using the `darts` python library. `darts` is similar to other libraries `scikit-learn` in Python or `tidymodels` in R that allow users to easily employ time series forecasting models without having to implement them. For model comparison, we rely on the `CRPS` library in Python to efficiently compute the continuous ranked probability scores (CRPS). When comparing the probabilistic forecasts, we express the CRPS score in error-orientation -- i.e., a lower CRPS indicates a smaller margin of error.

All the machine learning algorithms that are used in this study are based on neural network architectures. Neural networks have the property of being universal function approximators, so theoretically, it is possible that all of these models could exactly approximate the data generating process \cite{}. Yet, it is often true that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock, makridakis}. This performance gap can be due to a variety of reasons including overfitting and insufficient hyperparameter tuning \cite{adcock}. The discrepancy between theory and practice in function approximation with neural networks motivates research on how machine learning models perform in specific domains.

The 8 machine learning models that we compare take a variety of approaches with the design of their neural networks. We will not go into detail on how these different models work, but for those interested to learn more, we list the models and their references in table \ref{}. For readers who do not specialize in ML, an important concept to understand is that the ML models learn directly from the data and are not instilled with domain knowledge. This non-mechanisitic basis is at once very powerful as it does not restrict the models with misleading assumptions but is simultaneously limiting as these models are not readily interpretable and generally require more data than mechanistic analogs \cite{}. For the management of critical resources like water, this lack of interpretability could be a deterrent to the use of ML methods, and there may not be enough data to accomodate ML approaches \cite{}.

The ML models are configured to generate probabilistic forecasts. The neural network architectures that are used in this study are deterministic, so they are not able to directly represent probability distributions. We work around this limitation by performing quantile regression whereby the neural networks are trained to output quantiles at each time step in the forecast window. Forecasts are then generated by drawing samples according to these quantiles.

@@

\begin{table}[!htbp]
    \centering
    \caption{Mean CRPS and RMSE}
    \begin{subtable}{\textwidth}
        \centering
        \caption{Dissolved Oxygen}
        \begin{tabular}{lcc}
            \toprule
            Model & Mean CRPS & Mean RMSE \\
            \midrule
            Climatology & 0.62 & 0.92 \\
            Naive Persistence & & 1.55 \\
            AutoTheta & 0.67 & 1.02 \\
            BlockRNN & 0.52 & 0.80 \\
            DLinear & 0.52 & 0.81  \\
            NBEATS & 0.51 & 0.80 \\
            NLinear & 0.52 & 0.80 \\
            \textbf{NaiveEnsemble} & \textbf{0.47} & \textbf{0.74} \\
            RNN & 0.60 & 0.92 \\
            TCN & 0.62 & 0.98 \\
            TFT & 0.52 & 0.80 \\
            Transformer & 0.51 & 0.80 \\
            \bottomrule
        \end{tabular}
    \end{subtable}
    \begin{subtable}{\textwidth}
        \centering
        \caption{Water Temperature}
        \begin{tabular}{lcc}
            \toprule
            Model & Mean CRPS & Mean RMSE \\
            \midrule
            Climatology & 1.46 & 2.19 \\
            Naive Persistence & & 6.05 \\
            AutoTheta & 1.55 & 2.35 \\
            BlockRNN & 1.45 & 2.26 \\
            DLinear & 1.35 & 2.08 \\
            NBEATS & 1.36 & 2.10 \\
            NLinear & 1.33 & 2.06 \\
            \textbf{NaiveEnsemble} & \textbf{1.18} & \textbf{1.80} \\
            RNN & 1.32 & 2.03 \\
            TCN & 1.89 & 3.02 \\
            TFT & 1.19 & 1.86 \\
            Transformer & 1.32 & 2.05 \\
            \bottomrule
        \end{tabular}
    \end{subtable}
    \begin{subtable}{\textwidth}
        \centering
        \caption{Chlorophyll-a}
        \begin{tabular}{lcc}
            \toprule
            Model & Mean CRPS & Mean RMSE \\
            \midrule
            Climatology & 4.83 & 7.50 \\
            \textbf{Naive Persistence} & & \textbf{4.86} \\
            AutoTheta &  4.20 & 6.31 \\
            \textbf{BlockRNN} & \textbf{3.14} & 5.16 \\
            DLinear & 3.47 & 5.48 \\
            NBEATS & 3.78 & 5.95 \\
            NLinear & 3.67 & 5.63 \\
            NaiveEnsemble & 3.44 & 5.50 \\
            RNN & 4.15 & 6.40 \\
            TCN & 3.72 & 5.77 \\
            TFT & 3.67 & 5.87 \\
            Transformer & 4.02 & 6.35 \\
            \bottomrule
        \end{tabular}
    \end{subtable}
    \label{tab:mean_crps_rmse}
\end{table}

>>>>

There are many promising neural network architectures that have not yet been evaluated for water quality forecasting. Most applications of neural networks to limnological time series have focused on the long-short-term-memory (LSTM) network, a model that was published in 1997 \cite{lstm}. Over the nearly 30 years since LSTM's were introduced, researchers in computer science have built upon these earlier neural network architectures, introducing new models that achieve state of the art performance \cite{tft, tcn, blockrnn, NBEATS}. Many recent studies that use neural networks for water quality forecasting have not explored more contemporary neural network architectures. We will address this gap in the literature by comparing 8 neural network models including LSTM's as well as more recently published neural network models.

@@

We evaluated the forecasting performance of 12 different forecasting models on water temperature, dissolved oxygen and chlorophyll-A time series recorded by in-situ sensors at 34 freshwater sites across North America. These sites consist of Lakes, Non-wadeable Rivers and Wadeable Streams, subtypes classified by NEON. There is variability across sites in which target variables were observed. Additionally, maintenance issues led to gaps in the data which also varied across locations. Provided the lack of time series and large gaps at certain sites, we evaluated the forecast performance for water temperature, dissolved oxygen and chlorophyll-A at 33, 32 and 9 sites, respectively.

The imputation of missing data proved to be critical to the performance of the ML models. After experimenting with data filling methods that resulted in poor model performance, we developed an imputation method inspired by the climatology model. If the gap size was less than 5 days, then the gap was filled using a Gaussian Process Filter. For gaps over 5 days, missing data was estimated using the daily historical mean when this statistic is available; and, when there are no data collected for a day of the year, either the monthly median, quarterly median, previous observation or global median was used in this order of preference.

@@

where $\nabla$ is the difference operator, $Z_t$ is defined as a theta line and $X_t$ is the original data. When $\theta < 1$, the second differences are reduced from the data, yielding a Theta line that amplifies the long term trends of the data. For instance, if $\theta = 0$, the theta line will be a linear regression of the data. And if $\theta > 1$, the short term behavior of the data will be magnified in the Theta line. The Theta model generates a forecast by extrapolating the linear combination of two or more Theta lines. Although the Theta model is relatively simple, it has performed remarkably well in prominent forecasting competitions \cite{}. In the time since the Theta model was originally presented in 2000, new variations of the Theta model have been presented which have been shown to outperform the original Theta model \cite{}. Throughout this study, we use the StatsForecast AutoTheta model which selects the best performing model from a selection of variations of the Theta model.

@@

All the machine learning algorithms that are used in this study are based on neural network architectures. Neural networks have the property of being universal function approximators, so in theory, it is possible that all of these models could exactly approximate the data generating process \cite{}. Yet, it is often true that neural networks greatly underperform their function approximation capabilities in practice \cite{adcock, makridakis}. This performance gap can be due to a variety of reasons including overfitting and insufficient hyperparameter tuning \cite{adcock}. The discrepancy between theory and practice in function approximation with neural networks motivates research on how machine learning models perform in specific domains.

The 8 machine learning models that we compare take a variety of approaches with the design of their neural networks. We will not go into detail on how these different models work, but for those interested to learn more, we list the models and their references in table \ref{}. For readers who do not specialize in ML, an important concept to understand is that the neural network models learn directly from the data and are not instilled with domain knowledge. This non-mechanisitic basis is at once very powerful as it does not restrict the models with misleading assumptions, but neural networks are also limiting in that they are not readily interpretable and require more data than knowledge-guided methods \cite{justinread2019}. For the management of critical resources like water, the lack of interpretability could be a deterrent to the adoption of NN methods, and there may not be enough data for some water systems to accommodate ML approaches \cite{}.

The neural network models are configured to generate probabilistic forecasts. The neural network architectures that are used in this study map real-valued vectors to real-valued vectors, so they are not able to directly produce probabilistic forecasts. We work around this limitation by performing quantile regression whereby the neural networks are trained to output quantiles at each time step in the forecast window. Forecasts are then generated by drawing samples according to these quantiles.

For the 8 neural network models that we investigate, there are varying design choices made regarding the type of covariates that can be used. For this study, we employed 2 groups of models, one group used past covariates and the other used future covariates. For the models that accept past covariates (which included TCN, BlockRNN, NLinear, DLinear, Transformer and NBEATS), we used the other target variables recorded at that site as well as air temperature as covariates. For the models that only accept future covariates (which included TFT and LSTM), we used the day of the year as the covariate. The historical time series were split into a training and a validation set, whereby the models were trained on time series from 2020 to 2023 and validated at 12 30-day non-overlapping intervals in 2023.

We provide fully reproducible code for fitting, scoring and visualizing the forecasts. All the machine learning forecasts were generated using the `darts` python library \cite{}. `darts` is similar to other libraries like `scikit-learn` in Python or `tidymodels` in R in that the library allows users to employ a variety of time series forecasting models without having to implement them.

@@

We examine the forecast skill of 8 neural network models (LSTM, BlockRNN, NBEATS, NLinear, DLinear, TFT, TCN and Transformer), 1 statistical model (AutoTheta) and 2 null models (Naive Persistence and Climatology) on time series taken from the NEON Forecasting Challenge's Aquatics Theme. In addition to these individual models, we created an ensemble model that aggregates the forecasts taken from all the neural network models. Following recent work that has established that multi-model ensembles offer advantages over individual models for water temperature forecasting \cite{freya-mme}, we were inspired to investigate multi-model ensembles in this comparative study. For the neural network multi-model ensemble, all the NN models are represented equally, hence its name "Naive Ensemble".

In Table \ref{}, we present the mean CRPS and RMSE scores for the respective target variables. For Dissolved Oxygen (DO) and Water Temperature (WT), there are similar results for model performance: the neural network models outperform the AutoTheta, Naive Persistence and Climatology model with few exceptions; and the naive ensemble model is the best performing model with respect to both CRPS and RMSE. Yet, with Chlorophyll-A, there are much different results: while the neural network models still generally outperform the reference models in CRPS, the Naive Ensemble model is no longer the top performing model in either CRPS or RMSE. BlockRNN attains the best performance with respect to CRPS and the Naive Persistence model attains the best mean RMSE.

Intuition for why we see the results in the scoring tables can be gained by looking at individual forecasts shown in Figure \ref{}.

\begin{table}[!htbp]
    \centering
    \caption{Mean CRPS and RMSE}
    \begin{subtable}{\textwidth}
        \centering
        \caption{Dissolved Oxygen ($\text{mg} \, \text{L}^{-1}$)}
        \begin{tabular}{lcc}
            \toprule
            Model & Mean CRPS & Mean RMSE \\
            \midrule
            Climatology & 0.62 & 0.92 \\
            Naive Persistence & & 1.55 \\
            AutoTheta & 0.67 & 1.02 \\
            BlockRNN & 0.52 & 0.80 \\
            DLinear & 0.52 & 0.81  \\
            NBEATS & 0.51 & 0.80 \\
            NLinear & 0.52 & 0.80 \\
            \textbf{NaiveEnsemble} & \textbf{0.47} & \textbf{0.74} \\
            RNN & 0.60 & 0.92 \\
            TCN & 0.62 & 0.98 \\
            TFT & 0.52 & 0.80 \\
            Transformer & 0.51 & 0.80 \\
            \bottomrule
        \end{tabular}
    \end{subtable}
    \begin{subtable}{\textwidth}
        \centering
        \caption{Water Temperature ($^{\circ} \text{C}$)}
        \begin{tabular}{lcc}
            \toprule
            Model & Mean CRPS & Mean RMSE \\
            \midrule
            Climatology & 1.46 & 2.19 \\
            Naive Persistence & & 6.05 \\
            AutoTheta & 1.55 & 2.35 \\
            BlockRNN & 1.45 & 2.26 \\
            DLinear & 1.35 & 2.08 \\
            NBEATS & 1.36 & 2.10 \\
            NLinear & 1.33 & 2.06 \\
            \textbf{NaiveEnsemble} & \textbf{1.18} & \textbf{1.80} \\
            RNN & 1.32 & 2.03 \\
            TCN & 1.89 & 3.02 \\
            TFT & 1.19 & 1.86 \\
            Transformer & 1.32 & 2.05 \\
            \bottomrule
        \end{tabular}
    \end{subtable}
    \begin{subtable}{\textwidth}
        \centering
        \caption{Chlorophyll-a ($\text{mg} \, \text{L}^{-1}$)}
        \begin{tabular}{lcc}
            \toprule
            Model & Mean CRPS & Mean RMSE \\
            \midrule
            Climatology & 4.83 & 7.50 \\
            \textbf{Naive Persistence} & & \textbf{4.86} \\
            AutoTheta &  4.20 & 6.31 \\
            \textbf{BlockRNN} & \textbf{3.14} & 5.16 \\
            DLinear & 3.47 & 5.48 \\
            NBEATS & 3.78 & 5.95 \\
            NLinear & 3.67 & 5.63 \\
            NaiveEnsemble & 3.44 & 5.50 \\
            RNN & 4.15 & 6.40 \\
            TCN & 3.72 & 5.77 \\
            TFT & 3.67 & 5.87 \\
            Transformer & 4.02 & 6.35 \\
            \bottomrule
        \end{tabular}
    \end{subtable}
    \label{tab:mean_crps_rmse}
\end{table}
